SUPPLEMENTARY:\\A new approach to bias correction in high-throughput sequencing data

### 2011-11-22 seqbias-paper.tex
Added p_max/d_max plots, fixed ensemble 60/62 mixup, removed the use of the word 'classification'.

Towards that end, we propose training a Bayesian network classifier trained on
examples of foreground and background sequences. Though our goal is not
classification, by using the machinery of classification we can avoid a model
that is over-parametrized. Parameters that are not informative in
discriminating between foreground and background are not included in the model.
The Bayesian network can then be used to evaluate sequence probability, and thus
bias, at any genomic position. Figure \ref{fig:overview} gives a schematic overview
of the proposed model.

@@

To evaluate the efficacy of each of the bias correction methods considered, we
counted reads overlapping each gene, defining the gene by the union of every
transcript in release 62 of the Ensembl gene annotations, excluding UTRs, and
100 additional bases from both ends of the gene. Counts were then normalized by
dividing by the length of these genes. We then removed any genes with a read
count less than 10, or that did not correspond to a unique TaqMan probe.

@@

\begin{table}
\processtable{
    Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
    values.\label{tab:pcr}}{
    \centerline{
\begin{tabular*}{0.4\textwidth}{lrl}
\toprule
Method & Correlation & \\ \midrule
Unadjusted &  0.6665 & $\ast\ast$ \\
7mer &  0.6677 & $\ast\ast$ \\
GLM & 0.6854 & $\ast\ast$ \\
MART & 0.6921 & $\ast$ \\
BN & \textbf{0.7016}
\end{tabular*}}
}
{
We estimated the statistical significance of the improvement in correlation
using the BN method over the other methods using a simple boostrap procedure. A
bootstrap sample is formed by sampling, with replacement, 592 genes from the
original set of the same size. The correlation is then computed for this set,
using the adjusted count from each method. We repeated this procedure one
million times, and counted the number of times each of the competing methods
achieved a higher correlation than the BN method.  Those marked with a sigle
asterisk achieved a higher correlation fewer than 1000 times, resulting in a
p-value less than $10^{-3}$.  Those marked with two asterisks achieved a higher
correlation in none of the bootstrap samples, indicating a p-value less than
$10^{-6}$.
}
\end{table}

>>>>

Towards that end, we propose training a Bayesian network using examples of
foreground and background sequences. By training the model discriminatively and
penalizing model complexity, we can avoid a model that is over-parametrized,
excluding parameters that insufficiently informitive in discriminating between
foreground and background. The Bayesian network can then be used to evaluate
sequence probability, and thus bias, at any genomic position. Figure
\ref{fig:overview} gives a schematic overview of the proposed model.

@@

To evaluate the efficacy of each of the bias correction methods considered, we
counted reads overlapping each gene, defining the gene by the union of every
transcript in release 60 of the Ensembl gene annotations. Counts were then
normalized by dividing by the length of these genes. We then removed any genes
with a read count less than 10, or that did not correspond to a unique TaqMan
probe.

@@

\begin{table}
\processtable{
    Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
    values.\label{tab:pcr}}{
    \centerline{
\begin{tabular*}{0.4\textwidth}{lrl}
\toprule
Method & Correlation & \\ \midrule
Unadjusted &  0.6650 & $\ast\ast$ \\
7mer &  0.6680 & $\ast\ast$ \\
GLM & 0.6874 & $\ast\ast$ \\
MART & 0.6998 & $\ast$ \\
BN & \textbf{0.7086}
\end{tabular*}}
}
{
We estimated the statistical significance of the improvement in correlation
using the BN method over the other methods using a simple boostrap procedure. A
bootstrap sample is formed by sampling, with replacement, 648 genes from the
original set of the same size. The correlation is then computed for this set,
using the adjusted count from each method. We repeated this procedure one
million times, and counted the number of times each of the competing methods
achieved a higher correlation than the BN method.  Those marked with a sigle
asterisk achieved a higher correlation fewer than 1000 times, resulting in a
p-value less than $10^{-3}$.  Those marked with two asterisks achieved a higher
correlation in none of the bootstrap samples, indicating a p-value less than
$10^{-6}$.
}
\end{table}

### 2011-11-22 seqbias-supplement.tex
Added p_max/d_max plots, fixed ensemble 60/62 mixup, removed the use of the word 'classification'.

>>>>

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-p_max.pdf}
\end{center}
\caption{Sensitivity over values of the $p_{\text{max}}$ parameter,
controlling the maximum in-degree of any node in the directed graph. The $Y$ axis is
as in Figure \ref{fig:poissampstd}.}
\label{fig:pois_p_max}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-d_max.pdf}
\end{center}
\caption{Sensitivity over values of the $d_{\text{max}}$ parameter, controlling
the maximum distance any edge in the directed graph may traverse.  The $Y$ axis
is as in Figure \ref{fig:poissampstd}.}
\label{fig:pois_d_max}
\end{figure}

Figures \ref{fig:pois_p_max} and \ref{fig:pois_d_max} suggest that most
dependencies are between adjacent nucleotides, as increasing the maximum
in-degree ($p_{\text{max}}$), or maximum edge distance ($d_{\text{max}}$) has
little effect for values of at least 1.
